Enhanced Global PTM Discovery (G-PTM-D) with MetaMorpheus

### 2017-12-05 paper.tex
djfkdjf

The database augmentation step allowed 208 modification types, including PTMs, adducts, and chemical modifications.
See Supplementary Table 1 for a complete listing.

@@

A critical parameter for peptide identification is mass accuracy\citep{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systematic drift and miscalibration all limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised and fall into three general categories: external calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\citep{Olsen_2005}); and spectral calibration subsequent to the MS experiment (post-acquisition spectral calibration).
We developed a post-acquisition calibration procedure that builds upon the software lock mass concept\citep{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem MS spectra are compiled, and then used to recalibrate the spectra.
The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 20 ppm fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired FDR.
Step 2: The identifications from Step 1 provide a theoretical list of m/z peaks that should match closely to observed m/z peaks in the spectra file, and the discrepancies between the theoretical and observed peaks are used to inform a peak shifting procedure.
The calibration algorithm is described in detail in Supplementary Material 2.

@@

In a conventional search, only the isolation m/z value of 726.65 would be used to select a range of theoretical peptides for comparison.
Assuming correct deisotoping and charge state calculation, the mass 1447.74 Da would be searched.
However, all peptides in the range 725.4 to 727.9 Th are co-fragmented and a better solution would be to use all co-isolated precursor masses.
In MetaMorpheus, for each MS2 scan, the corresponding MS1 scan is deconvoluted and the complete list of precursors is gathered for the range surrounding the isolated precursor ($\pm 1.25$ Th).
Each deconvoluted mass is used to select a range of theoreticals to use for identification.
The result of this is that often, more than one peptide can be identified from a single MS2 scan.
This is extremely beneficial in both open mass and multi-notch searches because it reduces the number of the more difficult to interpret mass differences and substantially improves G-PTM-D (see below).

@@

The next step, database augmentation, outputs significantly different augmented databases because of the difference between accepted mass differences in the open and the multi-notch searches.
At this point, another advantage of the enhanced approach is clear; the overall search time dropped significantly, from 22.1 hours to 2.9 hours.

@@

We thank the members of the Smith lab for testing G-PTM-D and other MetaMorpheus functionality (Leah V. Schaffer, Anthony J. Cesnik, Rob Millikin, Lei Lu, Zach Rolfs, Sean Dai, Rachel Miller, Rachel Knoener), who contributed input to the improvement of mzLib, and MetaMorpheus. We thank the MetaMorpheus users outside of this lab who submitted useful comments on GitHub. This work was supported by the grant DBI-1458524 from the National Science Foundation.

>>>>

The database augmentation step allowed 208 modification types, including PTMs, adducts, and chemical modifications.
See Table S-1 for a complete listing.

@@

A critical parameter for peptide identification is mass accuracy\citep{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systematic drift and miscalibration all limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised and fall into three general categories: external calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\citep{Olsen_2005}); and spectral calibration subsequent to the MS experiment (post-acquisition spectral calibration).
We developed a post-acquisition calibration procedure that builds upon the software lock mass concept\citep{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem MS spectra are compiled, and then used to recalibrate the spectra.
The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 20 ppm fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired FDR.
Step 2: The identifications from Step 1 provide a theoretical list of m/z peaks that should match closely to observed m/z peaks in the spectra file, and the discrepancies between the theoretical and observed peaks are used to inform a peak shifting procedure.
The calibration algorithm is described in detail in Document S-2.

@@

In a conventional search, only the isolation m/z value of 726.65 would be used to select a range of theoretical peptides for comparison.
Assuming correct deisotoping and charge state calculation, the mass 2175.96 Da would be searched.
However, all peptides in the range 725.4 to 727.9 Th are co-fragmented and a better solution would be to use all co-isolated precursor masses.
In MetaMorpheus, for each MS2 scan, the corresponding MS1 scan is deconvoluted and the complete list of precursors is gathered for the range surrounding the isolated precursor ($\pm 1.25$ Th).
Each deconvoluted mass is used to select a range of theoreticals to use for identification.
The result of this is that often, more than one peptide can be identified from a single MS2 scan, as shown in Figure~\ref{fig:fig5-coIsolationSpectrum}.
This is extremely beneficial in both open mass and multi-notch searches because it reduces the number of the more difficult to interpret mass differences and substantially improves G-PTM-D (see below).

@@

Further improvements arise from using a multi-notch search to create the augmented database.
The enhanced G-PTM-D procedure employed a narrow 5 ppm tolerance around 208 mass difference notches (see Document S-2).
The resulting augmented database is greatly improved compared to using an open search for this step.
At this point, another advantage of the enhanced approach is clear; the overall search time dropped significantly from 22.1 hours to 2.9 hours, primarily due to the narrow tolerance multi-notch search versus the open search.

@@

We thank the members of the Smith lab (Leah V. Schaffer, Anthony J. Cesnik, Rob Millikin, Lei Lu, Zach Rolfs, Sean Dai, Rachel Miller, and Rachel Knoener) for testing G-PTM-D and contributing to the improvement of mzLib and MetaMorpheus.
We thank the MetaMorpheus users outside of this lab who submitted useful comments on GitHub.
This work was supported by the grant DBI-1458524 from the National Science Foundation.
